---
title: "Data Analysis using R"
author: '500666461'
date: "`r Sys.Date()`"
output: pdf_document
---

Loading required libraries *dplyr*, *stringr* and *ggplot2*

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
```

## Task 1

Reading the dataset into R

```{r}
cdrc_data <- read.csv("hh_churn_lsoa11_2023.csv")
str(cdrc_data) #extraneous
```

## Task 2

Examining the LSOA code W01000092 Menai (Bangor), and identifying the residential mobility indices which are greater than 0.5 and less than 0.8.

```{r}
cdrc_data_menai_df <- filter(cdrc_data, area == "W01000092")
cdrc_data_menai <- unlist(cdrc_data_menai_df[1, -1])

cdrc_data_menai_sub <- cdrc_data_menai[cdrc_data_menai > 0.5 & cdrc_data_menai < 0.8]
# printing the named vector
print(cdrc_data_menai_sub)
# printing the names of the vector
print(names(cdrc_data_menai_sub))
# printing the actual years
x <- str_replace(names(cdrc_data_menai_sub), "chn", "")
print(x)
```

## Task 3

Appending additional columns to the data

```{r}
# adds column one (averaged residential mobility for each region) as ARMI
cdrc_data$ARMI <- round(rowMeans(cdrc_data[, -c(1, 28)]), 3)
# adds column two (grouping regions into low, medium and high) as ARMIgrpd
cdrc_data$ARMIgrpd <- cut(cdrc_data$ARMI, breaks = c(0, 0.2, 0.5, 1), labels = c("Low", "Medium", "High"))
str(cdrc_data)
```

## Task 4

### First insight (with chart)

```{r}
avg_armi <- round(colMeans(cdrc_data[, 2:27]), 3)
years <- as.numeric(str_replace(names(avg_armi), "chn", ""))
ts_df <- data.frame(year = years, avg_rmi = avg_armi)

ggplot(ts_df, aes(x = year, y = avg_rmi)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  labs(x = "Year",
       y = "Average residential mobility index",
       title = "Trend of residential mobility index over time across regions") +
  theme_bw()
```

The chart in Figure 1 is a time series plot showing the change in residential mobility index (averaged across all the regions) over time from 1997 to 2022. The plot clearly shows a sharp downward trend in the residential mobility index over time. The downward tredn is also observed to be uniform, with the line approximately straight. The inference from the chart is that, on average, as years pass by, the residential mobility decreases. And this applies to all regions.

### Second insight (with insight)

```{r}
ggplot(cdrc_data, aes(x = chn1997, y = chn2022, color = ARMIgrpd)) +
  geom_point() +
  labs(x = "Residential mobility index in 1997",
       y = "Residential mobility index in 2022",
       color = "RMI level",
       title = "Correlation plot for the residential mobility index - year 2022 against 1997") +
  theme_bw()
```

The residential mobility indices in 2022 is plotted against the residential mobility indices for 1997, to detect correlation in the indices across the years.























